---
title: "ISL Chapter 7 Lab -- Moving Beyond Linearity"
output: html_notebook
---


# ISL Chapter 7 -- Moving Beyond Linearity

# Non-Linear Modeling

In this lab, we re-analyze the Wage data considered in the examples throughout this chapter, in order to illustrate the fact that many of the complex non-linear fitting procedures discussed can be easily implemented in R. We begin by loading the ISLR library, which contains the data

```{r}
library(ISLR)
head(Wage)
```

# Polynomial Regression and Step Functions

We first fit the model 


```{r}
fit <- lm(wage ~ poly(age, 4), data = Wage)
coef(summary(fit))
```

This syntax fits a linear model, using the lm() function, in order to predict wage using a fourth-degree polynomial in age: poly(age,4).

The poly command allows us to avoid having to write out a long ormula with powers of age. The function returns a matrix whose columns are a basis of *orthogonal polynomials*, which essentially means that each column is a linear combination of the variables age, age^2, age^3, age^4. 

However, we can also use poly() to obtain age, age^2, age^3, age^4 directly if we prefer. We can do this by using the raw=TRUE argument to the poly() function. Later we see that this does not affect the model in a meaningful way -- thought the choice of basis clearly affects the coefficient estimates, it does not affect the fitted values obtained

```{r}
fit2 <- lm(wage ~ poly(age, 4, raw = TRUE), data = Wage)
coef(summary(fit2))
```

We now create a grid of values for age at which we want predictions, and then call the generic predict() function, specifying that we want standard errors as well

```{r}
agelims <- range(age) # get min and max for age
age_grid <- seq(from = agelims[1], to = agelims[2]) # we want 1 prediction for each unique value of age --> 63 unique values
preds <- predict(fit, newdata = list(age = age_grid), se = TRUE)

se_bands <- cbind(preds$fit + 2 *preds$se.fit, preds$fit - 2 *preds$se.fit) # 95% CI (2 ~ 0.525 quantile for Standard Gaussian) it's an approximation (should prolly use student quantiles instead)
preds_or <- preds # copy for later
se_bands_or <- se_bands # copy for later

# plot the data and add the fit from the degree-4 polynomial:
plot(age, wage, xlim = agelims, cex = 0.5, col = "darkgrey")
title("Degree-4 Polynomial")
lines(age_grid, preds$fit, lwd = 2, col = "blue")
matlines(age_grid, se_bands, lwd = 1, col = "blue", lty = 3)
```

We mentioned earlier that whether or not an orthogonal set of basis functions is produced in the poly function will not affect the model obtained in a meaningful way. What do we mean by this? The fitted values obtained in either case are identical:

```{r}
preds2 <- predict(fit2, newdata = list(age = age_grid), se = TRUE)
max(abs(preds$fit - preds2$fit)) #maximum abs difference between predicted values --> very small value
```

When performing a poly regression we must decide on the degree of the polynomial to use. One way to do this is by using hypothesis tests.

We now fit models ranging from linear to a degree-5 polynomial and seek to determine the simplest model which is sufficient to explain the relationship between age and wage. 

We use the anova() function, which performs an analysis of variance (ANOVA, using an F-test) in order to test the null hypothesis that a model M1 is sufficient to explain the data against the alternative hypothesis that a more complex model M2 is required.

In order to use the anova() function, M1 and M2 must be *nested* models: the predictors in M1 must be a subset of the predictors in M2.

In this case we fit five different models and sequentially compare the simpler model to the more complex model:

```{r}
fit1 <- lm(wage ~ age, data = Wage)
fit2 <- lm(wage ~ poly(age,2), data = Wage)
fit3 <- lm(wage ~ poly(age,3), data = Wage)
fit4 <- lm(wage ~ poly(age,4), data = Wage)
fit5 <- lm(wage ~ poly(age,5), data = Wage)
anova(fit1, fit2, fit3, fit4, fit5)
```

The p-value comparing the linear M1 to the quadratic M2 is essentially zero, indicating that a linear fit is not sufficient. Similarly the p-value comparing the quadratic M2 to the cubic M3 is very low, so the quadratic fit is also insufficient. The p-value comparing the cubic and degree-4 polynomials, is approx 5%, while the degree-5 polynomial M5 seems unnecessary because the p-value is 0.37.

Hence, either a cubic or a quartic polynomial appear to provide a reasonable fit to the data, but lower- or higher-order models are not justified.

In this case, instead of using the anova() function, we could have obtained these p-values more succintly by exploiting the fact that poly() creates orthogonal polynomials:

```{r}
coef(summary(fit5))
```
Notice that the p-values are the same, and in fact the square of the t-statistics are equal to the F-statistics from the anova() function

However, the ANOVA method works whether or not we used orthogonal polynomials; it also works when we have other terms in the model as wel.

For example, we can use anova() to compare these three models:
```{r}
fit1 <- lm(wage ~ education + age, data = Wage)
fit2 <- lm(wage ~ education + poly(age,2), data = Wage)
fit3 <- lm(wage ~ education + poly(age,3), data = Wage)
fit4 <- lm(wage ~ education + poly(age,4), data = Wage)
anova(fit1, fit2, fit3, fit4)
```

As an alternative to using hyp. tests and anova, we could choose the polynomial degree using cross-validation. 


Next we consider the task of predicting whether an individual earns more tha 250k dollars per year. We proceed much as before, except that first we create the appropriate response vector, and then apply the glm() function using family = "binomial" in order to fit a polynomial logistic regression model.

```{r}
fit <- glm(I(wage > 250) ~ poly(age, 4), data = Wage, family = binomial)
```

Note that we again use the wrapper I() to create this binary response variable on the fly. The expression wage > 250 evaluates to a logical variable containing TRUEs and FALSEs, which glm() coerces to binary by setting TRUEs to 1s and FALSEs to 0s.

Once again we make predictions using the predict function:
```{r}
preds <- predict(fit, newdata = list(age = age_grid) ,se = TRUE)
```

However, calculating the confidence intervals is slightly more involved than in the linear regression case. The default prediction type for a glm() model is type = "link", which is what we use here. This means we get predictions for the logit (log odds) = X x Beta. The standard errors given are also of this form. In order to get confidence intervals for Pr(Y = 1 | K) we use the transformation:


Pr(Y = 1 | X) = exp(X x Beta) / (1 + exp(X x Beta))

```{r}
pfit <- exp(preds$fit) / ( 1 + exp(preds$fit)) # probabilities Pr(Y = 1 | X)
se_bands_logit <- cbind(preds$fit + 2 * preds$se.fit, preds$fit - 2 * preds$se.fit) # 95% CI for log odds
se_bands <- exp(se_bands_logit) / (1 + exp(se_bands_logit)) # 95% CI for probs
```
Note that we could have directly computed the probs by selecting the type = "response" option in the predict() function.
```{r}
preds <- predict(fit, newdata = list(age = age_grid), type = "response", se = TRUE)
```

However, the corresponding CI would not have been sensible because we would end up with negative probabilities.
Finally the right hand plot:
```{r}
par(mfrow = c(1,2), mar = c(4.5,4.5,1,1), oma = c(0,0,4,0))
# left hand plot
plot(age, wage, xlim = agelims, cex = 0.5, col = "darkgrey")
title("Degree-4 Polynomial", outer = TRUE)
lines(age_grid, preds_or$fit, lwd = 2, col = "blue")
matlines(age_grid, se_bands_or, lwd = 1, col = "blue", lty = 3)



# right hand plot
{plot(age, I(wage>250), xlim = agelims, type = "n", ylim = c(0, .2))
points(jitter(age), I((wage > 250)/5), cex =.5, pch = "|", col = "darkgrey")
lines(age_grid, pfit, lwd = 2, col = "blue")
matlines(age_grid, se_bands, lwd = 1, col = "blue", lty = 3)}

```

Here the mar and oma arguments to par allow us to control the margins of the plot, and the title function creates a figure title that spans both subplots.

We have drawn the age values corresponding to the obs with wage values above 250 as grey marks on the top of the plot, and those with wage values below 250 are shown as gray marks on the bottom of the plot. We used the jitter() function to jitter the age values a bit so  that obs with the same age value do not cover each other up. This is often called a rug plot.

In order to fit a step function, we use the cut() function.

```{r}
table(cut(age,4)) 
fit <- lm(wage ~ cut(age, 4), data = Wage)
coef(summary(fit))

```
Here cut() automatically picked the cutpoints at 33.5, 49 and 64.5 years of age. We could also have specified our own cutpoints directly using the breaks option. The function cut() returns an ordered categorical variable; the lm() function then creates a set of dummy variables for use in the regression. The age < 33.5 category is left out, so the intercept coefficient of 94k dollars can be interpreted as the average salary for those under 33.5 years of age, and the other coefficients can be interpreted as the average additional salary for those in the other age groups. We can produce predictions and plots just as we didd in the case of the polynomial fit.

# Splines

In order to fit regression splines in R, we use the splines library. Regression splines can be fit by constructing an appropriate matrix of basis functions. The bs() function generates the entire matrix of basis functions for splines with the specified set of knots. By default, cubic splines are produced. Fitting wage to age using a regression spline is simple:
```{r}
library(splines)
fit <- lm(wage ~ bs(age, knots = c(25, 40, 60)), data = Wage)
preds <- predict(fit, newdata = list(age = age_grid), se = TRUE)
{plot(age, wage, col = "gray")
lines(age_grid, preds$fit, lwd = 2)
lines(age_grid, preds$fit + 2 * preds$se, lty = "dashed")
lines(age_grid, preds$fit - 2 * preds$se, lty = "dashed")
}
```

Here we have specified knots at ages 25, 40, and 60. This produces a spline with six basic functions. (Recall that a cubic spline with three knots has seven degrees of freedom; these degrees of freedom are used up by an intercept, plus six basis functions.)
We could also use the df option to produce a spline with knots at uniform quantiles of the data:
```{r}
dim(bs(age, knots = c(25, 40, 60)))
dim(bs(age, df = 6))
attr(bs(age, df = 6), "knots")
```
In this case R chooses knots at ages 33.8, 42.0, and 51.0, which correspond to the 25th, 50th, and 75th percentiles of age. The function bs() also has a degree argument, so we can fit splines of any degree, rather than the default degree of 3 (which yields a cubic spline).

In order to instead fit a natural spline, we use the ns() function. Here we fit a natural spline with four degrees of freedom.
```{r}
fit2 <- lm(wage ~ ns(age, df = 4), data = Wage)
preds2 < predict(fit2, newdata = list(age_grid), se = T)
{plot(age, wage, col = "gray")
lines(age_grid, preds$fit, lwd = 2)
lines(age_grid, preds2$fit, col = "red", lwd = 2)
}
```
As with the bs() function, we could instead specify the knots directly using the knots option.
In order to fit a smoothing spline, we use the smooth.spline() function:

```{r}
fit <- smooth.spline(age, wage, df = 16)
fit2 <- smooth.spline(age, wage, cv = TRUE)
fit2$df
{plot(age, wage, xlim = agelims, cex = .5, col = "darkgrey")
title("Smoothing Spline")
lines(fit, col = "red", lwd = 2)
lines(fit2, col = "blue", lwd = 2)
legend("topright", legend = c("16 DF", "6.8 DF"), col = c("red", "blue"), lty = 1, lwd = 2, cex = .8)
}


```
Notice that in the first call to smooth.spline(), we specified df = 16. The function then determines which value of lambda leads to 16 degrees of freedom. In the second call to smooth.spline(), we select the smoothness level by crossvalidation; this results in a value of lambda that yields 6.8 degrees of freedom. 


In order to perform local regression, we use the loess() function.

```{r}
fit <- loess(wage ~ age, span = .2, data = Wage)
fit2 <- loess(wage ~ age, span = .5, data = Wage)

{plot(age, wage, xlim = agelims, cex = .5, col = "darkgrey")
title("Local Regression")
lines(age_grid, predict(fit, data.frame(age = age_grid)), col = "red", lwd = 2)
lines(age_grid, predict(fit2, data.frame(age = age_grid)), col = "blue", lwd = 2)
legend("topright", legend = c("Span = 0.2", "Span = 0.5"), col = c("red", "blue"), lty = 1, lwd = 2, cex = .8)
}

```


Here we have performed local linear regression using spans of 0.2 and 0.5: that is, each neighborhood consists of 20% or 50% of the observations. The larger the span, the smoother the fit. 

# GAMS

We now fit a GAM to predict wage using natural spline functions of year and age, treating education as a qualitative predictor. Since this is just a big linear regression model using an appropriate choice of basis functions, we can simply do this using the lm() function.

```{r}
gam1 <- lm(wage ~ ns(year, 4) + ns(age, 5) + education, data = Wage)
```
We now fit the model using smoothing splines rather than natural splines. In order to fit more general sorts of GAMs, using smoothing splines or other components that cannot be expressed in terms of basis functions and then fit using least squares regression, we will need to use the gam library in R.

The s() function, which is part of the gam library(), is used to indicate that we would like to use a smoothing spline. We specify that the function of year should have 4 degrees of freedom, and that the function of age will have 5 degrees of freedom. Since education is qualitative, we leave it as is, and it is converted into four dummy variables. We use the gam() function in order to fit a GAM using these components. All of the terms are fit simultaneosly, taking each other into account to explain the response.
```{r}
library(gam)
gam_m3 <- gam(wage ~ s(year, 4) + s(age, 5) + education, data = Wage)
par(mfrow = c(1,3))
plot(gam_m3, se = TRUE, col = "blue")
plot.Gam(gam1, se = TRUE, col = "red")
```
Notice here we hade to used plot.Gam() rather than the generic plot().

In these plots the functon of year looks rather linear. We can perform a series of ANOVA tests in order to determine which of these three models is best: a GAM that excludes year(M1), a GAM that uses a linear function of year (M2), or a GAM that uses a spline function of year (M3).
```{r}
gam_m1 <- gam(wage ~ s(age,5) + education, data = Wage)
gam_m2 <- gam(wage ~ year + s(age,5) + education, data = Wage)
anova(gam_m1, gam_m2, gam_m3, test = "F")

```

We find that there is compelling evidence that a GAM with a linear function of year is better than a GAM that does not include year at all (p-value = 0.00014).
However, there is no evidence that a non-linear function of year is needed. In other words, based on the results of this ANOVA, M2 is preferred.
The summary() function produces a summary of the gam fit.
```{r}
summary(gam_m3)
```

The p-values for year and age correspond to a null hypothesis of a linear relationship versus the alternative of a non-linear relationship. The large p-value for year reinforces our conclusion from the ANOVA test that a linear function is adequate for this term. However, there is very clear evidence that a non-linear term is required for age.
We can make predictions from gam objects, just like from lm objects, using the predict() method for the class gam.

```{r}
par(mfrow = c(1,3))
preds <- predict(gam_m2, newdata = Wage)
# we can also use local regression fits as building blocks in a GAM, suing the lo() function
gam_lo <- gam(wage ~ s(year, df = 4) + lo(age, span = 0.7) + education, data = Wage)
plot.Gam(gam_lo, se = TRUE, col = "green")
```
Here we have used local regression for the age term, with a span of 0.7. We can also use the lo() function to create interactions before calling the gam() function. For Example
```{r}
gam_lo_i <- gam(wage ~ lo(year, age, span = 0.5) + education, data = Wage)
```
fits a two-term model, in which the first term is an interaction between year and age, fit by a local regression surface. We can plot the resulting two-dimensional surface if we first install the akima package

```{r}
#library(akima)
#plot(gam_lo_i)
```
In order to fit a logistic regression GAM, we once again use the I() function in constructing the binary response variable, and set family = binomial
```{r}
gam_lr <- gam(I(wage > 250) ~ year + s(age, df = 5) + education, family = binomial, data = Wage)
par(mfrow = c(1,3))
plot(gam_lr, se = TRUE, col = "green")
```
It is east to see that there are no high earners in the <HS category:
```{r}
table(education, I(wage > 250))
```
Hencem we fit a logistic regression GAM using all but this category. This provides more sensible results.
```{r}
gam_lr <- gam( I(wage > 250) ~ year + s(age, df = 5) + education, family = binomial, data = Wage, subset = (education !="1. < HS Grad"))
plot(gam_lr, se = TRUE, col = "green")
```

