---
title: "ISL Chapter 4 Lab -- Classification"
output: html_notebook
---


# ISL Chapter 4 -- Classification


The Stock Market Data

Smarked dataset:

percentage returns for the S&P 500 stock index over 1250 days from the beginning of 2001 until the end of 2005.
For each date we have recorded: 

Lag1 through Lag5 -- the percentage returns for each of the five previous trading days
Volume -- the number of shares traded on the previous day, in billions
Today -- the percentage return on the date in question
Direction -- whether the market was Up or Down on this day

We will try to predict the Direction of the market on a certain day using info of the past


```{r}
library(ISLR)
library(magrittr)
library(dplyr)
head(Smarket)
dim(Smarket)
summary(Smarket)
```

Check correlation between our predictor (cannot compute it for our target variable since it's categorical)

```{r}
# I used pipe functionality from magrittr and select from dplyr
market_cor <- Smarket %>% select(-Direction) %>% cor()
market_cor
#apply(market_cor,1,function(x)ifelse(x >0.1, TRUE, FALSE))
```
The only sizable correlation is between Year and Volume --> each year the number of shares traded increases
```{r}
time_vs_volume <- ggplot(data = Smarket, aes(x = as.numeric(row.names(Smarket)), y = Volume))
time_vs_volume + geom_point(color = "#CC0000") + ylab("Shares traded -- billions") + xlab("Index") + theme_bw()
attach(Smarket)
plot(Volume)
head(row.names(Smarket))

```

# Logistic Regression

Let's fit a Logistic Regression model to predict Direction using Lag1 -- Lag5 and Volume.
The glm() function fits generalized linear models, a class of models that includes logistic regression.
To specify that we want a logistic regression we pass to it the argument family = binomial

```{r}
log_reg1 <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume, data = Smarket, family = binomial)
summary(log_reg1)
```

The smallest p-value (Pr(>|z|)) is associated to the predictor Lag1. The negative predictor suggests that if the market had a positive return yesterday, then it is less likely to go up today. However, at a vaue of 0.15, the p-value is still relatively large, and so there is no clear evidence of a real association between Lag1 and Direction.

We use the coef() function in order to access just the coefficients for this fitted model. We can also use summary() function to access particular aspects of the fitted model, such as the p-values for the coefficients:

```{r}
coef(log_reg1) # just the estimates
summary(log_reg1)$coef  # the full table
summary(log_reg1)$coef[,4]  #the p-values
```

The predict() function can be used to predict the probability that the market will go up, given values of the predictors.
The type = "response" option tells R to output probabilities of the form P(Y = 1| X), as opposed to other information such as the logit.
If no data set is suplied to the predict() function, then the probabilities are computed for the training data that was used to fit the logistic regression model.
Here we have printed only the first ten probabilities.
We know that these values correspond to the probability of the market going up, rather than down, because the contrasts() function indicates that R has created a dummy variable with a 1 for Up:
```{r}
log_reg1_probs <- predict(log_reg1, type = "response")
log_reg1_probs[1:10]
contrasts(Direction)
```

In order to make a prediction as to whether the market will go up or down on a particular day, we must convert these predicted probabilities into class labels, Up or Down. The following two commands create a vector of class predictions based on whether the probability of a market increased is greater than or less than 0.5.

Given these predictions, the table() function can be used to produce a confusion matrix in order to determine how many observations were correctly or incorrectly classified
```{r}
log_reg1_pred <- rep("Down", 1250) # initialize vector w/ length = # of obs, value --> Down
log_reg1_pred[log_reg1_probs > 0.5] = "Up" # change to up the element in the vector that have and index = to the index of elements of the vector of probabilies that have a value > 0.5  -- confusing explanation ^^'
table(log_reg1_pred, Direction)
(507+145) / 1250 # 
mean(log_reg1_pred==Direction) # number of obs predicted == observed

```
The diagonal elements of the confusion matrix indicate correct predictions, while the off-diagonals represent incorrect predictions.

Hence our model correctly predicted that the market would go up on 507 days and that it would go down on 145 days, for a total of 507 + 145 = 652 correct predictions.

The mean() function can be used to compute the fraction of days for which the prediction was correct. In this case, logistic regression correctly predicted the market 52.2% of the time.

At first glance, it appears that Log Reg is working slightly better than random guessing.
However this result is miselading because we trained and tested the model on the same dataset. In other words 100 - 52.2 = 47.8 % is the training error rate. The training error rate is often overly optimisting -- it tends to underestimate the test error rate. In order to better assess the accuracy of the logistic regression model in this setting, we can fit the model using part of the data, adn then examine how well it predicts the held out data. 

This will yield a more realistic error rate, in the sense that in practice we will be interested in our model's performance not on the data that we used to fit the model, but rather on days in the future for which the market's movements are unknown.


We will create a vector of observations from 2001 through 2004 to be used to fit(train) our model. <-- TRAINING DATA
We will use the traned model to predict observations from 2005 <-- TESTING DATA

```{r}
train <- (Year < 2005) # we can use just the column names, e.g. Year, because we attached their dataframe, Smarket, previously. 
Smarket_2005 <- Smarket[!train,] # select just rows for which train is FALSE
dim(Smarket_2005)
Direction_2005 <- Direction[!train] # same for the observed results --> this is the "truth" for the 2005 observations
```

We now fit a logistic regression model using only the subset of the observations that correspond to dates before 2005, using the subset argument.
We then obtain the predicted probabilities of the stock market going up for each of the days in our test set.

```{r}
log_reg2 <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume, data = Smarket, family = binomial, subset = train)
log_reg2_probs <- predict(log_reg2,Smarket_2005, type = "response") # now we specify the observations to predict, our test dataset
```

Finally, we compute the predictions for 2005 and compare them to the actual movements of the market over that time period.

```{r}
log_reg2_preds <- rep("Down", 252)
log_reg2_preds[log_reg2_probs > 0.5] = "Up"

table(log_reg2_preds)
mean(log_reg2_preds == Direction_2005) # obs correctly predicted
mean(log_reg2_preds != Direction_2005) # obs wrongly predicted

```


The TEST ERROR RATE is 52%, higher than the training error rate (which was 47.8 %), Our binary classifier is slightly worse than tossing a fair coin!

Of course this result is not surprising --> it's simply not feasible to predict the stock market using the previous days returns.

We recall that the logistic regression model had very underwhelming p-values associated with all of the predictors, and that the smallest p-value, though not very small, corresponded to Lag1.

Perhaps by removing the variables that appear not to be helpful in predicting Direction, we can obtain a more effective model.
After all, using predictors that have no relationship with the response tends to cause a deterioration in the test error rate (since such predictors cause an increase in variance without a corresponding decrease in bias), and so removing such predictors may in turn yield an improvement.

Below we have refit the logistic regression using just Lag1 and Lag2, which seemed to have the highest predictive power in the original logistic regression model


```{r}
log_reg3 <- glm(Direction ~ Lag1 + Lag2, data = Smarket, family = binomial, subset = train)
log_reg3_probs <- predict(log_reg3,Smarket_2005, type = "response")
log_reg3_preds <- rep("Down", 252)
log_reg3_preds[log_reg3_probs > 0.5] = "Up"

table(log_reg3_preds)
mean(log_reg3_preds == Direction_2005) # obs correctly predicted
mean(log_reg3_preds != Direction_2005) # obs wrongly predicted
```

# Linear Discriminant Analysis

We will perform LDA on the same Smarket dataset.
In R we fit an LDA model using the lda() function, which is part of the MASS library. Notice that the syntax for the lda() function is identical to that of lm(), and to that of glm() except for the absence of the family option. We fit the model using only the observations before 2005.

```{r}
library(MASS)
lda1 <- lda(Direction ~ Lag1 + Lag2, data = Smarket, subset = train)
lda1
plot(lda1)

```

the LDA output indicates that the priors for Down and Up are 0.492 and 0.508 respectively. In other words, 49.2% of the training observations correspond to days during which the market went down. 
It also provides the group means; these are the average of each predictor within each class, and are used by LDA as estimates of mu k.

These suggest that there is a tendency for the previous 2 days' returns to be negative on days when the market increases, and a tendency for the previous days' returns to be positive on days when the market declines.

The coefficient of linear discriminants output provides the linear combination of Lag1 and Lag2 that are used to form the LDA decision rule.

In other words, these are the multipliers of the elements of X = x in the equation (4.19; pag 143). 
If -0.642 x Lag1 - 0.513 x Lag2 is large, then the LDA classifier will predict a market increase, and if it is small, then the LDA classifier will predict a market decline.


The plot() function produces plots of the linear discriminants, obtained by computing -0.642 x Lag1 - 0.514 x Lag2 for each of the training observations.
We can see that the two distributions are pretty similar --> it will be hard to classify correctly.

the predict() function returns a list with three elements. 
The first element, class, contains LDA's predictions about the movement of the market. 
The second element, posterior, is a matrix whose kth column contains the posterior probability that the corresponding observation belongs to the kth class, computed from (4.10 pag 139 -- Bayes' Theorem). 
Finally, x contains the linear discriminants, described earlier.

```{r}
lda1_pred <- predict(lda1, Smarket_2005)
names(lda1_pred)
lda1_class <- lda1_pred$class
table(lda1_class, Direction_2005)
```

LDA and logistic regression predictions are almost identical.

Applying a 50% threshold to the posterior probabilities allows us to recreate the predictions contained in lda1_pred$class

```{r}
# notice that the posterior probability output by the model corresponds to the probability that the market will decrease
# therefore:
sum(lda1_pred$posterior[,1] >= 0.5) # observations predicted Down
sum(lda1_pred$posterior[,1] < 0.5) # observations predicted Up
```

If we wanted to use a posterior probability other than 50% in order to make predictions, then we could easily do so. For instance, suppose that we wish to predict a market decrease only if we are very certain that the market will indeed decrease on that day -- say, if the posterior probability is at least 90%:

```{r}
sum(lda1_pred$posterior[,1] > 0.90) # observations predicted Down w/ theshold at 90% posterior probability
```

No days in 2005 meet that threshold! In fact, the greatest posterior probability of decrease in all of 2005 was 52.05%.


# Quadratic Discriminant Analysis

We will now fit a QDA model to the Smarket data. QDA is implemented in R using the qda() function, which is also part of the MASS library. The syntax is identical to that of lda().

```{r}
qda1 <- qda(Direction ~ Lag1 + Lag2, data = Smarket, subset = train)
qda1
```

The output contains the group means. But it does not contain the coefficients of the linear discriminants, because the QDA classifier involves a quadratic, rather than a linear, function of the predictors. 
The predict() function works in exactly the same fashion as for LDA.

```{r}
qda1_class <- predict(qda1, Smarket_2005)$class
table(qda1_class)
mean(qda1_class == Direction_2005) # percentage of correct predictions

```

Interestingly, the QDA predictions are accurate almost 60% of the time, even though the 2005 data was not used to fit the model. 
This level of accuracy is quite impressive for stock market data, which is known to be quite hard to model accurately. This suggests that the quadratic form assumed by QDA may capture the true relationship more accurately than the linear forms assumed by LDA and logistic regression.

# K-Nearest Neighboors

We will now perform KNN using the knn() function, which is part of the class library. This function works rather differently from the other model-fitting functions that we have encountered thus far. 

KNN doesn't make an assumption on the form of the relationship between predictor and target --> non-parametric approach (more flexible)

Rather than a two-step approach in which we first fit the model and then we use the model to make predictions, knn() forms predictions using a single command. The function requires four inputs:

1. A matrix containing the predictors associated with the training data, labeled train.X below.

2. A matrix containing the predictors associated with the data for which we wish to make predictions, labeled test.X below.

3. A vector containing the class labels for the training observations, labeled train.Direction below

4. A value for K, the number of nearest neighbors to be used by the classifier.

We use the cbind() fucntion, short for column bind, to bind the Lag1 and Lag2 variables together into two matrices, one for the training set and the other for the test set.

```{r}
library(class)
# we can call Smarket's columns just by their name because we attached the dataset earlier
# train is a boolean vector which is TRUE for the indexes corresponding of years from 2001 to 2004 and FALSe for the indexes corresponding to year 2005
train.X <- cbind(Lag1, Lag2)[train,] # predictors matrix -- training
test.X <- cbind(Lag1, Lag2)[!train,] # predictors matrix -- testing
train.Direction <- Direction[train] # labels
```

Now the knn() function can be used to predict the market's movement for the dates in 2005. We set a random seed before we apply knn() because if several obs are tied as nearest neighbors, the R will (pseudo)randomly break the tie. Therefore, a seed must be set in order to ensure reproducibility of results.

```{r}
set.seed(1)
knn1_pred <- knn(train.X, test.X, train.Direction, k = 1)
table(knn1_pred, Direction_2005)
(83 + 43) / 252
```

The results using K = 1 are not very good, since only 50% of the obs are correctly predicted. Of course, it may be that K = 1 results in an overly flexible fit to the training data. Below we repeat the analysis using K = 3 and k = 5:

```{r}

knn3_pred <- knn(train.X, test.X, train.Direction, k = 3)
table(knn3_pred, Direction_2005)
mean(knn3_pred == Direction_2005)
  
knn5_pred <- knn(train.X, test.X, train.Direction, k = 5)
table(knn5_pred, Direction_2005)
mean(knn5_pred == Direction_2005)
```

The results have improved slightly for k = 3. But increasing K further turns out to provide no further improvements, quite the contrary.
It appears that, for this data, QDA provides the best results of the methods that we have examined so far.

# An Application to Caravan Insurance Data

Finally, we will apply the KNN approach to the Caravan data set, which is part of the ISLR library. 

This data set includes 85 predictors that measure demographic characteristics for 5822 individuals.
The response variable is Purchase, which indicates whehter or not a given individual purchases a caravan insurance policy.
In this data set, only 6% of people purchased caravan insurance

```{r}
dim(Caravan)
attach(Caravan)
summary(Purchase)
```

Because the KNN predicts the class of a given test observation by identifying the observations that are nearest to it, the scale of the variables matters. Any variables that are on a large scale will have a much larger effect on the distance between the observations, and hence on the KNN classifier, than variables that are on a small scale. 

For instance, imagine a data set that contains two variables, salary and age (measured in dollars and years, respectively). As far as KNN is concerned, a difference of 1000$ in salary is enourmous compared to a difference of 50 years in age. Consequently, salary will drive the KNN classification results, and age will have almost no effect. This is contrary to our intuition that a salary difference of 1000 dollars is quite small compared to an age difference of 50 years. Furthermore, the importance of scale to the KNN classifier leads to another issue: if we measured salary in Japanese yen, or if we measured age in minutes, then we'd get quite different classification results from what we get if these two variables are measured in dollars and years.

A good way to handle this problem is to standardize the data so that all variables are given a mean of zero and a standard deviation of one. Then all variables will be on a comparable scale. The scale() function does just this. In standardizing the data, we exclude column 86, because that is the qualitative Purchase variable, our target.

```{r}
standardized.X <- scale(Caravan[, -86])
# variance for the first two features of raw data
var(Caravan[,1])
var(Caravan[,2])
# variance for the first two features of standardized data
var(standardized.X[,1])
var(standardized.X[,2])
```

Now every column(feature) of standardized.X has a standard deviation of one and a mean of zero.

We now split the observations into a test set, containing the first 1000 observations, and a training set, containing the remaining observations.
We fit a KNN model on the training data using K = 1, and evaluate its performance on the test data:

```{r}
test <- 1:1000 # the first 1000 persons will be our test set 
train.X <- standardized.X[-test,]
test.X <- standardized.X[test,]
train.Y <- Purchase[-test]
test.Y <- Purchase[test]
set.seed(1)
knncar1_pred <- knn(train.X, test.X, train.Y, k = 1)
mean(test.Y != knncar1_pred) #percentage of wrong predictions using our knn model
mean(test.Y != "No") #percentage of wrong predictions if we just predicted every label as "No"
```

The KNN error rate on the 1000 test obs is just under 12%. At first glance this may appear to be fairly good. However, since only 6% of customers purchased insuracne, we could get the error rate down 6% by always predictong No regardless of the values of the predictors.


Suppose that there is some non-trivial cost to trying to sell insurance to a given individual. Perhaps a salesperson must visit each potential customer.
If the company tries to sell insurance to a random selection of customers, then the success rate will be only 6%, which may be far too low fiven the cost involved. Instead, the company would like to try to sell insurance only to customers who are likely to buy it. 

So the overall error rate is not of interest.

Instead, the fraction of individuals that are correctly predicted to buy insurance is of interest.

It turns out that KNN with K = 1 does far better than random guessing among the customers that are predicted to buy insurance. 
Among 77 such customers, 9, or 11.7% actualy do purchase insurance. This is double the rate that one would obtain from random guessing.

```{r}
table(knncar1_pred, test.Y)
9/(68 + 9)
```


Using K = 3 the success rate increases to 19%, and with K = 5 the rate is 26.7%. This is over four times tha rate that results from random guessing.

It appears that KNN is finding some real patterns in a diffcult data set!


```{r}
knncar3_pred <- knn(train.X, test.X, train.Y, k = 3)
table(knncar3_pred, test.Y)
5 / (21 + 5)
  
knncar5_pred <- knn(train.X, test.X, train.Y, k = 5)
table(knncar5_pred, test.Y)
4 / (11 + 4)
```

As a comparison, we can also fit a logistic regression model to the data.
If we use 0.5 as the predicted probability cut-off for the classifier, then we have a problem: only seven of the test observations are predicted to purchase insurance. Even worse, we are wrong about all of these.

However, we are not required to use a cut-off of 0.5. If we instead predict a purchase any time the predicted probability of purchase exceeds 0.25, we get much better results: we predict that 33 people will purchase insurance, and we are correct for about 33% of these people. This is over five times better than random guessing.

```{r}

glmcar <- glm(Purchase~., data = Caravan, family = binomial, subset = -test)
glmcar_probs <- predict(glmcar, Caravan[test,], type = "response")
glmcar_pred <- rep("No", 1000)
glmcar_pred[glmcar_probs > 0.5] <- "Yes"
table(glmcar_pred, test.Y)

# try to change cut-off 

glmcar_pred <- rep("No", 1000)
glmcar_pred[glmcar_probs > 0.25] <- "Yes"
table(glmcar_pred, test.Y)
11/(22 + 11)

```
